Introducing statistics with \\texttt{intRo}

### 2014-12-12 student-paper-competition-2014/paper.Rmd
edits to paper

The widespread adoption of \texttt{R} as a tool for statistical analysis has undoubtedly been a great development for the scientific community. However, using \texttt{R} in most cases still requires a basic knowledge of programming concepts which may be unfamiliar to the introductory statistics student. This issue may explain why introductory courses often utilize software applications such as \texttt{JMP}, even though instructors would tend to use a tool like \texttt{R} in their own work. Still, signficant compromises must be made when using software like \texttt{JMP}, including dealing with software licenses, unsupported desktop platforms, and non-ideal user interface decisions.

In teaching STAT 226 at Iowa State University, we witnessed profound struggles by students attempting to practice the introductory concepts discussed in class. Scrimshaw notes in his manuscript that "open-ended packages, like any others, may create obstacles to learning simply through their lack of user-friendliness in the sheer mechanics of operating them, rather than any intrinsic difficulty in the content...." [@scrimshaw2001computers] In our own experience teaching, students' struggles were often directly related to the use of the software and not any sort of fundamental misunderstanding of the material, in agreement with Scrimshaw's finding.

Other software has recently been written in an attempt to spur student interest in introductory statistics and R programming. DataCamp's [@datacamp] courses are user-friendly ways to learning basic \texttt{R} programming and data analysis techniques. \texttt{Swirl} [@swirl] is a similar interactive tool to make learning \texttt{R} more fun by learning it within \texttt{R} itself. Project MOSAIC [@mosaic] has created a suite of tools to simplify the teaching of statistics in the form of an R package. The primary goal of DataCamp and Swirl is to teach \texttt{R} programming, rather than facilitate the learning of introductory statistics. Project MOSAIC's goal is to faciliate this learning, but using the package requires a knowledge of R programming that the introductory student may not have.

Upon the release of \texttt{RStudio}'s \texttt{Shiny} [@shiny] it became easy for an R-based analysis to be converted to an interactive web application. This in turn, led to the creation of an introductory statistics application which we call \texttt{intRo}, available at http://www.intro-stats.com. \texttt{intRo} offers a number of key advantages over traditional statistics software, which will be discussed at length in this text.

@@

In order for an application to be useful to students in their first statistics course, it can't hog the spotlight. Statistics as a subject should shine through as a fun and interesting topic to study, rather than being bogged down by difficult to learn software. With this in mind, in the development of \texttt{intRo} we focused on aspects of the user interface (UI) and output that make it easy to pick up without extensive training. We used large, easy to click icons in the page header to help students find exactly what they need easily. With this easy to use framework, we also made the functionality available in \texttt{intRo} the minimal necessary for an introductory statistics course. Additional functionality is nice, but if it clutters the UI beyond the point of usability, it will frustrate an introductory student. Figure \ref{fig:user_experience} illustrates the simple steps a student takes to generate a result in \texttt{intRo}, and view the code producing the result.

@@

\begin{figure*}[ht!]
\centering
\includegraphics[width=\linewidth]{images/ui_annotate.pdf}
\caption{The five elements that comprise the \texttt{intRo} application: 1) top navigation, 2) side navigation, 3) options panel, 4) results panel, and 5) code panel.}
\label{fig:ui}
\end{figure*}

 1. **Top Navigation** - The top navigation bar includes two sets of buttons. The left-aligned buttons are informational buttons. The first is a link to \texttt{intRo}. The second is a link to this documentation page. The third is a link to the GitHub repository where the code for \texttt{intRo} is housed. The final button is a link to our websites, which contain contact information if there are any questions or comments. The right-aligned buttons are \texttt{intRo} utilities. The first is a button to toggle the visibility of the code panel (5). The last is a button to print the stored module results, and the associated code (if visible).
 2. **Side Navigation** - The side navigation panel includes the modules that perform a specific data analysis task.
 3. **Options Panel** - The options panel includes module-specific options which the user can use to customize their results.
 4. **Results Panel** - The results pane displays the result of the selected module and options.
 5. **Code Panel** - The code panel displays the \texttt{R} code used to generate the results from the student's \texttt{intRo} session. The code panel is shown by default to facilitate an interest in coding, but can be hidden by clicking the code toggle button in the Top Navigation bar.

There are multiple modules within \texttt{intRo}, which perform specific data analysis tasks that employ an easy to use point-and-click interface. Each module is placed within a higher level category to further help students navigate the tool. The default modules included in \texttt{intRo} are split into three higher level categories - data, summaries, and statistics - and more modules can be easy added by an instructor, as detailed in section \ref{sec:extensible}. The default modules support uploading and downloading a dataset, transforming variables, graphical and numerical summaries, simple linear regression, contingency tables, and T-tests.

@@

## Exciting
In general, introduction to statistics is a class undergraduate students dread. It's hard, the language is unfamiliar, and the software is difficult to use. With \texttt{intRo}, we hope to flip this paradigm. \texttt{intRo} is a fun, easy to use tool that's available on the web, and is also usable on mobile devices. As a web-based application, this tool is immediately more familiar to students than a desktop application. With interactive plots from \texttt{ggvis} [@ggvis], \texttt{intRo} brings a `wow'-factor to learning introductory statistics.

@@

The process by which \texttt{intRo} is dynamically created upon running the application through module sourcing is detailed in figure \ref{fig:app_creation_modules}. The key driver to populating \texttt{server.R} and \texttt{ui.R} is the file \texttt{modules.txt}, which defines the placement of each module. The interface is then created with the following statement.

@@

# Reactive programming

@@

\texttt{intRo} leverages reactive programming techniques as data is manipulated by the user, and this is carried throughout the entire session. In a typical \texttt{Shiny} app, users interact with inputs, which in turn yield different results. Within \texttt{intRo}, the users are able to interact and manipulate the data underlying the entire application. This posed many challenges in the creation of \texttt{intRo}, namely timely save points according to the user's workflow, and reactive updating of variable lists tied to inputs across the entire application. Because the user may experiment with different configurations or select different variables, we did not want to store all actions taken in the intRo session. Rather, each module includes a button allowing the user to explicitly store the output visible in the results panel into the R script. This way, output is only stored when the user is satisfied, and the resulting output is not cluttered with unnecessary information.

In the creation of \texttt{intRo} we walked a fine line between giving the user flexibility and having realistic usability. At the same time, \texttt{intRo} was created as a consumer of another package, \texttt{Shiny}, in which we as developers were the beneficiaries of another team of developers' decision to balance flexibility and usability. For a tangible example, consider the graphical summaries module. We only allow variables of a type consistent with the selected plot to be displayed. This is a conscious decision that limits an \texttt{intRo} user's flexibility, while maximizing the usability (by minimizing crashes) of the application. On the flip side of this, \texttt{Shiny} allows much higher flexibility. For instance, the entire application (including user interface) is created dynamically upon load, based on the modules currently housed within \texttt{intRo}, as detailed in section \ref{sec:extensible}. However, \texttt{Shiny} does have limits on its flexibility based on the designers decisions for usability. One current example the slider element. This element allows for fixed width steps from its minimum to its maximum. The JavaScript library being utilized in \texttt{Shiny} allows for arbitrary function calls to to generate these steps, however they must be written in plain JavaScript. This is an example of a decision made by the developers of \texttt{Shiny} to limit functionality in favor of usability of their package.

@@

In addition to the current functionality, there are some practical improvements in the works that will make \texttt{intRo} more useful to both students and instructors. In particular, we have begun development on an \texttt{R} package which will allow \texttt{intRo} modules to be created automatically from user written \texttt{R} code. This package will generate the necessary file structure to allow the module's incorporation into \texttt{intRo} as well as translate user code to \texttt{intRo} compatible code as populates the necessary files. This will vastly improve \texttt{intRo}'s flexibility and allow it to be used in a wider range of curricula, including more advanced statistics courses. Furthermore, we would like to utilize the expanded interactive capabilities of \texttt{ggvis} in order to make \texttt{intRo's} plots more engaging to students. One way to do this would be implementing linked plots, in which interactions with one plot are reflected in other plots that illustrate the same data. This would be particularly useful in the regression module so that students could explore observations with high influence and leverage.

Additionally,  \texttt{intRo} will be trialed in an introductory statistics course at Iowa State University in Spring 2015. This will allow us to assess its usefulness relative to software used in the past, as well as guage areas for improvement. Furthermore, we can determine the effectiveness of code printing on generating excitement from the students about programming in \texttt{R}.

>>>>

The widespread adoption of \texttt{R} as a tool for statistical analysis has undoubtedly beenan important development for the scientific community. However, using \texttt{R} in most cases still requires a basic knowledge of programming concepts which may be unfamiliar to the introductory statistics student. This issue may explain why introductory courses often utilize software applications such as \texttt{JMP}, even though instructors would tend to use a tool like \texttt{R} in their own work. Still, signficant compromises must be made when using software like \texttt{JMP}, including dealing with software licenses, unsupported desktop platforms, and non-ideal user interface decisions.

In teaching Introduction to Business Statistics at Iowa State University, we witnessed profound struggles by students attempting to practice the introductory concepts discussed in class using current software. Scrimshaw notes in his manuscript that "open-ended packages, like any others, may create obstacles to learning simply through their lack of user-friendliness in the sheer mechanics of operating them, rather than any intrinsic difficulty in the content...." [-@scrimshaw2001computers] In our own experience teaching, students' struggles were often directly related to the use of the software and not any sort of fundamental misunderstanding of the material, in agreement with Scrimshaw's finding.

Multiple software packages have recently been written in an attempt to spur student interest in introductory statistics and R programming. DataCamp's [@datacamp] courses are user-friendly ways to learning basic \texttt{R} programming and data analysis techniques. \texttt{Swirl} [@swirl] is a similar interactive tool to make learning \texttt{R} more fun by learning it within \texttt{R} itself. Project MOSAIC [@mosaic] has created a suite of tools to simplify the teaching of statistics in the form of an R package. The primary goal of DataCamp and Swirl is to teach \texttt{R} programming, rather than facilitate the learning of introductory statistics. Project MOSAIC's goal is to faciliate this learning, but using the package requires a knowledge of R programming that the introductory student may not have.

Upon the release of \texttt{RStudio}'s \texttt{Shiny} [@shiny] it became easier for an R-based analysis to be converted to an interactive web application. This in turn led us to create an introductory statistics application which we call \texttt{intRo}, available at http://www.intro-stats.com. \texttt{intRo} offers a number of key advantages over traditional statistics software, including ease of use and an aim to foster student interest. Additionally, \texttt{intRo} stands apart from new tols in that it is a supplement to an existing class, fully usable by a beginning statistics student. These advantages will be discussed at length in this text.

@@

In order for an application to be useful to students in their first statistics course, it can't hog the spotlight. Statistics as a subject should shine through as a fun and interesting topic to study, rather than being bogged down by difficult to learn software. With this in mind, in the development of \texttt{intRo} we focused on aspects of the user interface (UI) and output that make it easy to pick up without extensive training. We used large, easy to click icons in the page header to help students find exactly what they need easily. With this easy to use framework, we also made the functionality available in \texttt{intRo} the minimal necessary for an introductory statistics course. Additional functionality is nice, but if it clutters the UI beyond the point of usability, it will frustrate an introductory student. Figure \ref{fig:user_experience} illustrates the simple steps a student takes to generate a result in \texttt{intRo}, view the code producing the result, and save the results for later.

@@

\begin{figure*}[ht!]
\centering
\includegraphics[width=\linewidth]{images/ui_annotate.pdf}
\caption{The five elements that comprise the \texttt{intRo} application: 1) top navigation, 2) side navigation, 3) options panel, 4) results panel, and 5) code panel.}
\label{fig:ui}
\end{figure*}
 1. **Top Navigation** - The top navigation bar includes two sets of buttons. The left-aligned buttons are informational buttons. The first is a link to \texttt{intRo}. The second is a link to this documentation page. The third is a link to the GitHub repository where the code for \texttt{intRo} is housed. The final button is a link to our websites, which contain contact information if there are any questions or comments. The right-aligned buttons are \texttt{intRo} utilities. The first is a button to toggle the visibility of the code panel (5). The last is a button to print the stored module results, and the associated code (if visible).
 2. **Side Navigation** - The side navigation panel includes the modules that perform a specific data analysis task.
 3. **Options Panel** - The options panel includes module-specific options which the user can use to customize their results.
 4. **Results Panel** - The results pane displays the result of the selected module and options.
 5. **Code Panel** - The code panel displays the \texttt{R} code used to generate the results from the student's \texttt{intRo} session. The code panel is shown by default to facilitate an interest in coding, but can be hidden by clicking the code toggle button in the Top Navigation bar.

There are seven default modules within \texttt{intRo}, which perform specific data analysis tasks that employ an easy to use point-and-click interface. Each module is placed within a higher level category to further help students navigate the tool. The default modules included in \texttt{intRo} are split into three higher level categories - data, summaries, and statistics - and more modules can easily be added by an instructor, as detailed in section \ref{sec:extensible}. The default modules support uploading and downloading a dataset, transforming variables, graphical and numerical summaries, simple linear regression, contingency tables, and T-tests.

@@

## Exciting
In general, introduction to statistics is a class undergraduate students dread. It's hard, the vocabulary is unfamiliar, and the software is difficult to use. With \texttt{intRo}, we hope to flip this paradigm. \texttt{intRo} is a fun, easy to use tool that's available on the web, and is also usable on mobile devices. As a web-based application, this tool is immediately more familiar to students than a desktop application. With interactive plots from \texttt{ggvis} [@ggvis], \texttt{intRo} brings a `wow'-factor to learning introductory statistics.

@@

The process by which \texttt{intRo} is dynamically created upon running the application through module sourcing is detailed in figure \ref{fig:app_creation_modules}. The key driver to populating \texttt{server.R} and \texttt{ui.R} is the navigation file \texttt{modules.txt}, which defines the placement of each module. The interface is then created with the following statement.

@@

# Reactive design choices

@@

\texttt{intRo} leverages the reactive programming nature of \texttt{Shiny}, and as such is designed around the idea of user input cascading through the entire application. In a typical \texttt{Shiny} application, users interact with inputs that act as parameters to function, which in turn yield different results. Within \texttt{intRo}, the users are able to interact with and manipulate the data underlying the entire application. This posed many challenges in the creation of \texttt{intRo} and drove design decisions, namely timely save points according to the user's workflow, and reactive updating of variable lists tied to inputs across the entire application. Because the user may experiment with different configurations or select different variables, we did not want to store all actions taken in the intRo session. Rather, each module includes a button allowing the user to explicitly store the output visible in the results panel into the R script. This way, output is only stored when the user is satisfied, and the resulting output is not cluttered with unnecessary information.

In the creation of \texttt{intRo} we walked a fine line between giving the user flexibility and having realistic usability. At the same time, \texttt{intRo} was created as a consumer of another package, \texttt{Shiny}, in which we as developers were the beneficiaries of another team of developers' decision to balance flexibility and usability. For a tangible example, consider the graphical summaries module. We only allow variables of a type consistent with the selected plot to be displayed. This is a conscious decision that limits an \texttt{intRo} user's flexibility, while maximizing the usability (by minimizing crashes) of the application. On the flip side of this, \texttt{Shiny} allows much higher flexibility. For instance, the entire application (including user interface) is created dynamically upon load, based on the modules currently housed within \texttt{intRo}, as detailed in section \ref{sec:extensible}. However, \texttt{Shiny} does have limits on its flexibility based on the designers decisions for usability. One current example is the slider element. This element allows for fixed width steps from its minimum to its maximum. The JavaScript library being utilized in \texttt{Shiny} allows for arbitrary function calls to to generate these steps, however they must be written in plain JavaScript. This is an example of a decision made by the developers of \texttt{Shiny} to limit functionality in favor of usability of their package.

@@

In addition to the current functionality, there are some practical improvements in the works that will make \texttt{intRo} more useful to both students and instructors. In particular, we have begun development on an \texttt{R} package which will allow \texttt{intRo} modules to be created automatically from user written \texttt{R} code. This package will generate the necessary file structure to allow the module's incorporation into \texttt{intRo} as well as translate user code to \texttt{intRo} compatible code and populate the necessary files. This will vastly improve \texttt{intRo}'s flexibility and allow it to be used in a wider range of curricula, including more advanced statistics courses. Additionally, we would like to utilize the expanded interactive capabilities of \texttt{ggvis} in order to make \texttt{intRo's} plots more engaging to students. One way to do this would be implementing linked plots, in which interactions with one plot are reflected in other plots that illustrate the same data. This would be particularly useful in the regression module so that students could explore observations with high influence and leverage.

\texttt{intRo} will be trialed in an introductory statistics course at Iowa State University in Spring 2015. This will allow us to assess its usefulness relative to software used in the past, as well as guage areas for improvement. Furthermore, we can determine the effectiveness of code printing on generating excitement from the students about programming in \texttt{R}.
